---
title: "BATS_Niskins_KLv1"
author: "Krista Longnecker"
date: "`April 3, 2025`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Open up the libraries (Don't need all of these, but I did not go through the script to prune them)


```{r load libraries, message=FALSE, warning=FALSE}
rm(list = ls()) #start with a clean workspace (better for testing)

library(data.table) # you need to install this package from CRAN
library(dplyr)
library(readr) #this is part of the tidyverse package
library(here) #install this (once, install.packages("here))
#library(tidyverse) #amazing 
library(lubridate) #great for handling date data
library(zoo) #for interpolations
library(oce) #for integrations (package that will do oceanographic calculations)
library(ggplot2)
library(gridExtra)

library(reshape2)
```

STEP 1: Organizing BATS Bottle data downloaded from BATS web site.\
BATS data are available here: https://bios.asu.edu/bats/bats-data (as of 3 April 2025). The file with all the bottle information is : `bats_bottle.txt`


```{r GetFiles, message=FALSE, warning=FALSE}

rm(list=ls()) # clears the working environment to the right
options(stringsAsFactors=FALSE) # we want to avoid R creating factors when importing the data 

NameOfFile <- "BATS_Niskins.2025.04.03.RData"
#let's make this a little more uniform to make it easier to swap computers
here::here()

file.path <- "bats_bottle_KLdownload.txt"

# Load BATS data file into R
# This will skip over the first 59 lines from the file, up through the header
# BATS uses the first rows of the data file to set up the variable information.
# KL downloaded new version of file, let's use that one as that will be a better
# representative of what will get used in the future.
raw <- read.table(file.path,header = F, skip = 59)

```

STEP 2: This section puts the headers on columns

```{r STEP 2 - Add Headers to columns, message=FALSE, warning=FALSE}
# The BATS data file has a complicated format due to a combination of comma and tab-separation
# The header was removed above when file was loaded due to the commas which confuses R
# KL setting this up so the header is read in (avoid errors when changing variable names)

justHeaders <- read_lines(file.path, skip = 58, n_max = 1)
tt <- gsub(" ","",justHeaders) #remove the extra spaces
ttt <- gsub("\t","",tt) #remove the tabs
cleanedHeader <- strsplit(ttt,",") #this is now comma-separated, divide based on ","
header <- unlist(cleanedHeader) #result is a list, unlist so I can use it
rm(justHeaders,tt,ttt,cleanedHeader) #clean up variables no longer needed
colnames(raw) <- header #puts column names into header

# Change the name of our dataframe and remove missing data values and replace with NAs
data <- raw #defining a new object that uses raw and calls it a new "object" so as not to wreck the original object i.e. raw
data[data == -999] <- NA #anywhere in this object data where value is -999 we are going to call it NA
#str(data)

##use this opportunity to pull out month and date (using package lubridate)
data[,'month'] <- month(ymd(data[,'yyyymmdd']))
data[,'year'] <- year(ymd(data[,'yyyymmdd']))

rm(raw) #housecleaning

```


Remove stations more than 50 nm away from BATS (KL note 3 April 2025 --> this is off for now)

``` {r spatialMask, eval = FALSE}
#require points to be within some distance of BATS
lat = 31 + 40/60;
lon = (64 + 10/60)
BATS  <- data.frame(lat,lon)
HS <- data.frame(lat = 32.1667, lon =64.5000)

allPoints <- as.data.frame(data)

df <- as.data.frame(data)

#use geodDist from the R package oce to calculate distances from BATS
df$distance <- geodDist(df$lonW,df$latN,BATS$lon,BATS$lat)

howFar = 18.52 * 5  #remember: 10 nm ~ 18.52 km

#only keep data within the circle defined (check that definition)
data <- as.data.frame(df) %>%
  filter(distance < howFar) 

ggplot(data = allPoints, aes(x = lonW, y = latN),color = "year") + 
  geom_point(color = "gray") +
  geom_point(data = data,aes(x=lonW,y=latN),color = "blue") +
  geom_point(data = BATS,aes(x = lon, y = lat),color = 'red') +
  geom_point(data = HS,aes(x = lon, y = lat),color = 'magenta') +
  ggtitle("magenta = Hydrostation S; red = BATS; blue is stations within 50 nm")

#housecleaning
rm(BATS,HS,lat,lon,howFar)
  
```


STEP 3: The Id has the cruise type/number/cast/bottle information as one number --> split that up.

```{r splitID}
# break up ID to provide cruise, cast, and bottle ID
#data$cruise.type <- substr(data$Id, 1,1) #not doing this any more, use 5-digit cruise details
data$cruise <- substr(data$Id, 1,5) 
data$cast <- as.integer(substr(data$Id, 6,8))
data$bottle <- as.integer(substr(data$Id, 9,10))

##limit cruise years by year? (e.g., cruises with DON/TN data start in 1994)
#data <- data[data$yyyymmdd > 19931231,]

#reorganize a little to make the output easier to use
data <- data %>% relocate(cruise,.after=Id)
data <- data %>% relocate(cast,.after=cruise)
data <- data %>% relocate(bottle,.after=cast)


```

Export the results as a CSV file

```{r exportResult, warning=FALSE}

#export the results
write.csv(data,"BATSallTime_Niskins.csv", row.names=F) #cannot use this anymore, too messy

```

Also export the units so that people have that information (as a separate text file)

```{r unitInformation,eval = TRUE}
#This section will export the units (from the BATS file) as a text file

##Pull the units from the BATS text file (in the header).
#make a data frame with the header information and the units, easier to keep this separate
unitDetails <- as.data.frame(header)
getUnits <- read_lines(file.path,n_max = 58)

#start with idx <- 2 to skip over 'id'
for (idx in 2:length(header)){
  t <- pmatch(header[idx],getUnits) #t will be the row with the unit information
  a <- strsplit(getUnits[t] ,"=")
  unitDetails[idx,'details'] <- unlist(a)[2]
  #tidy up
  rm(t,a)
}
#tidy up
rm(getUnits,idx)

#The use of an equal sign for the Niskin/GoFlo flag causes a problem since 
#it will result in an unequal number of columns; I have otherwise split the 
#unit information based on the = sign; manually fix this
idx <- unitDetails$header %in% 'QF'
unitDetails[idx,'details'] <- "(-3 = suspect, 1=unverified, 2= verified/acceptable)"
rm(idx)

#before exporting the units trim & add to list as needed
idx <- unitDetails$header %in% colnames(data)
unitDetails <- unitDetails[idx,]
rm(idx)

#now add the new units...using their variable name as details for now
toAdd <- as.data.frame(setdiff(colnames(data),unitDetails$header))
colnames(toAdd) <- 'header'
toAdd$details <- toAdd$header
unitDetails <- rbind(unitDetails,toAdd)
rm(toAdd)

write.table(unitDetails,'BATSallTime_Niskins_unitDetails.txt',row.names=FALSE)

```


``` {r saveDetails}
sessionInfo()

save.image(NameOfFile)

```